---
title: "data311a1"
output: html_document
date: "2023-10-03"
---


## R Markdown

##q 3:

```{r}

##q 3a:

possum <- read.csv("~/data311lab1/possum.csv")

str(possum)

##q 3b:

possum <- possum[complete.cases(possum), ]

str(possum)

nrow(possum)

##q 3c:

set.seed(62197256)
n <- nrow(possum)
indicies <- sample(1:n, n)
train_size <- round(0.6*n)
test_size <- n - train_size
train = indicies[1:train_size]
test = indicies[(train_size+1):(train_size+test_size)]
training_data <- possum[train, ]
testing_data <- possum[test, ]

##q 3d:

colnames(training_data)

pairs(training_data[6:15])

hdlngth <- training_data$hdlngth

modelAge <- lm(hdlngth ~ training_data$age)
summary(modelAge)
##0.2219
modelSkull <- lm(hdlngth ~ training_data$skullw)
summary(modelSkull)
##0.5944
modelTot <- lm(hdlngth ~ training_data$totlngth)
summary(modelTot)
##0.4349
modelTail <- lm(hdlngth ~ training_data$taill)
summary(modelTail)
##0.02909
modelFoot <- lm(hdlngth ~ training_data$footlgth)
summary(modelFoot)
##0.2281
modelEar <- lm(hdlngth ~ training_data$earconch)
summary(modelEar)
##0.09552
modelEye <- lm(hdlngth ~ training_data$eye)
summary(modelEye)
##0.1444
modelChes <- lm(hdlngth ~ training_data$chest)
summary(modelChes)
##0.3987
modelBelly <- lm(hdlngth ~ training_data$belly)
summary(modelBelly)
##0.3208



modelSite <- lm(hdlngth ~ training_data$site)
summary(modelSite)
##0.1187
modelPop <- lm(hdlngth ~ training_data$Pop)
summary(modelPop)
##0.0456
modelSex <- lm(hdlngth ~ training_data$sex)
summary(modelSex)
##0.0111


possum_slr <- lm(hdlngth ~ training_data$skullw)
lm(hdlngth ~ training_data$skullw)
summary(possum_slr)

```


## the highest r^2 value is skull, with a width of 0.5944

```{r}

##q 3 e&f:

possum_mlr <- lm(hdlngth ~ training_data$taill + training_data$Pop)
lm(hdlngth ~ training_data$taill + training_data$Pop)
summary(possum_mlr)

```

##it seems that possums from victoria have larger heads as the est coeficient for vic is 2.3226
##which is quite large in magnitude & is positive - so it suggests possums from vic have bigger heads

```{r}

##q 3g:

possum_interact <- lm(hdlngth ~ training_data$taill * training_data$Pop)
lm(hdlngth ~ training_data$taill * training_data$Pop)
summary(possum_interact)

```

##the relationship seemed to be adequately capured by possum_mlr, so fitting with the interaction term didn't seem necesary

##q 3h:

## y = ( 0.6056 * a ) + ( 2.3226 * b )+ 68.8664
## where a = taill & b = Pop

```{r}

##q 3 i&j:

##q4
possum_slr <- lm(hdlngth ~ training_data$skullw)
lm(hdlngth ~ training_data$skullw)
s_slr <- summary(possum_slr)

p <- predict.lm(possum_slr, testing_data)[1:40]
mse_slr <- mean((testing_data$hdlngth - p) ^2)
mse_slr

##19.36403

##q5
possum_mlr <- lm(hdlngth ~ training_data$taill + training_data$Pop)
lm(hdlngth ~ training_data$taill + training_data$Pop)
summary(possum_mlr)

q <- predict.lm(possum_mlr, testing_data)[1:40]
mse_mlr <- mean((testing_data$hdlngth - q) ^2)
mse_mlr

##13.45257

##q7
possum_interact <- lm(hdlngth ~ training_data$taill * training_data$Pop)
lm(hdlngth ~ training_data$taill * training_data$Pop)
summary(possum_interact)

r <- predict.lm(possum_interact, testing_data)[1:40]
mse_itrct <- mean((testing_data$hdlngth - r) ^2)
mse_itrct

##13.73708

##t <- c( c(mse_slr, "mse_slr"), c(mse_mlr, "mse_mlr"), c(mse_itrct, "mse_itrct") )

##t <- c( c(mse_slr, mse_mlr, mse_itrct), c("mse_slr", "mse_mlr", "mse_itrct"))
##t <- table(t)
##rownames(t) = c("mse_slr", "mse_mlr", "mse_itrct")
##t

```

##table:

##    mse_itrct:          mse_mlr:            mse_slr: 
##  13.7370828794185    13.4525724013838    19.3640343694998

##I would deem the MLR (multi linear regression) to be the best
##because the main issue I would like to optimize against is minimizing MSE
##so the MLR with the lowest MSE is the best

##q 4:

##q4 setup:

```{r}

##setup:

possum <- read.csv("~/data311lab1/possum.csv")
str(possum)

possum <- possum[complete.cases(possum), ]
str(possum)
nrow(possum)

set.seed(62197256)
n <- nrow(possum)
indicies <- sample(1:n, n)
train_size <- round(0.6*n)
test_size <- n - train_size
train = indicies[1:train_size]
test = indicies[(train_size+1):(train_size+test_size)]
training_data <- possum[train, ]
testing_data <- possum[test, ]

colnames(training_data)
pairs(training_data[6:15])
hdlngth <- training_data$hdlngth

##library(class)
library(FNN)

length(training_data$totlngth)
length(training_data$hdlngth)

####training_data[,'class']=factor(training_data[,'class'])
####testing_data[,'class']=factor(testing_data[,'class'])

##knn1(training_data$totlngth, training_data$hdlngth, 2)

##knn1(training_data[1:40], testing_data, 2)

####knn(train=training_data$hdlngth[0:40], test=testing_data$hdlngth[0:40], cl=training_data$hdlngth[0:40], k=2)

```

##q 4a:

```{r}

train <- training_data$hdlngth
YYY <- training_data$totlngth
k2 <- FNN::knn.reg(train, test=NULL, y=YYY, k=2)

```

##q 4b:

```{r}

summary(k2)
q <- residuals(k2)
k2_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)

k2_mse

```

##q 4c:

```{r}

train <- training_data$hdlngth
YYY <- training_data$totlngth

```

```{r}

k3 <- FNN::knn.reg(train, test=NULL, y=YYY, k=3)
q <- residuals(k3)
k3_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k3_mse

k4 <- FNN::knn.reg(train, test=NULL, y=YYY, k=4)
q <- residuals(k4)
k4_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k4_mse

k5 <- FNN::knn.reg(train, test=NULL, y=YYY, k=5)
q <- residuals(k5)
k5_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k5_mse

k6 <- FNN::knn.reg(train, test=NULL, y=YYY, k=6)
q <- residuals(k6)
k6_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k6_mse

k7 <- FNN::knn.reg(train, test=NULL, y=YYY, k=7)
q <- residuals(k7)
k7_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k7_mse

k8 <- FNN::knn.reg(train, test=NULL, y=YYY, k=8)
q <- residuals(k8)
k8_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k8_mse

k9 <- FNN::knn.reg(train, test=NULL, y=YYY, k=9)
q <- residuals(k9)
k9_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k9_mse

k10 <- FNN::knn.reg(train, test=NULL, y=YYY, k=10)
q <- residuals(k10)
k10_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k10_mse

k11 <- FNN::knn.reg(train, test=NULL, y=YYY, k=11)
q <- residuals(k11)
k11_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k11_mse

k12 <- FNN::knn.reg(train, test=NULL, y=YYY, k=12)
q <- residuals(k12)
k12_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k12_mse

k13 <- FNN::knn.reg(train, test=NULL, y=YYY, k=13)
q <- residuals(k13)
k13_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k13_mse

k14 <- FNN::knn.reg(train, test=NULL, y=YYY, k=14)
q <- residuals(k14)
k14_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k14_mse

k15 <- FNN::knn.reg(train, test=NULL, y=YYY, k=15)
q <- residuals(k15)
k15_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k15_mse

k16 <- FNN::knn.reg(train, test=NULL, y=YYY, k=16)
q <- residuals(k16)
k16_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k16_mse

k17 <- FNN::knn.reg(train, test=NULL, y=YYY, k=17)
q <- residuals(k17)
k17_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k17_mse

k18 <- FNN::knn.reg(train, test=NULL, y=YYY, k=18)
q <- residuals(k18)
k18_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k18_mse

k19 <- FNN::knn.reg(train, test=NULL, y=YYY, k=19)
q <- residuals(k19)
k19_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k19_mse

k20 <- FNN::knn.reg(train, test=NULL, y=YYY, k=20)
q <- residuals(k20)
k20_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
k20_mse

```

```{r}

k2_mse
k3_mse
k4_mse
k5_mse
k6_mse
k7_mse
k8_mse
k9_mse
k10_mse
k11_mse
k12_mse
k13_mse
k14_mse
k15_mse
k16_mse
k17_mse
k18_mse
k19_mse
k20_mse

```

```{r}

k4_mse

```

##k=4 has the smallest MSE at 8693.271 

##q 4d:

```{r}

train <- training_data$hdlngth
##YYY <- c(training_data$pop, training_data$totlngth)
YYY <- training_data$taill

```

```{r}

kFinal <- FNN::knn.reg(train, test=NULL, y=YYY, k=2)
q <- residuals(kFinal)
kF_mse <- mean((testing_data$hdlngth - q[1:40]) ^2)
kF_mse

```

## the lowest MSE I could get was 8717.037, so the final model is k=4